Research log

### 2025-08-31 docs/ExploratoryResults.Rmd
Psych Science revisions

classoption       : "man"
output            : bookdown::gitbook
bibliography: references.bib
---

@@

groundhog.library(c(
  'png',
  'grid',
  'ggplot2',
  'svglite',
  'xtable',
  'papaja',
  'tidyverse',
  'broom',
  'cowplot',
  'MESS', # for AUCs
  'lsr', # for effect sizes
  'pwr', # for power calculations
  'brms', # for mixed effects modeling
  'BayesFactor', # for Bayesian t test
  'jsonlite', # parsing data from sort_trial
  'caret', #for cross validation
  'ggrepel' #for word scatterplots
), '2022-12-01')

@@

p<- E2.click_df %>%
  filter(!(subj_id %in% E2.lucky_pretenders))%>%
  group_by(subj_id,test_part, genuine_first) %>%
  summarise(sd_num_clicks=sd(num_clicks)) %>%
  group_by(test_part)%>%
  summarise(
    mean=mean(sd_num_clicks),
    se=se(sd_num_clicks)
    ) %>%
  ggplot(aes(x=test_part,
             color=test_part,
             y=mean))+
  geom_point(size=10)+
  geom_errorbar(aes(ymin=mean-se,ymax=mean+se),color='black')+
  scale_color_manual(values=c("#69b3a2", "#404080")) +

@@

E2.debrief_labeled <- read.csv('..\\experiments\\Battleships2\\data\\debrief.csv',na.strings=c(""," ","NA")) %>%
  dplyr::select(-subj_id) %>%
  merge(E2.debrief <- E2.df %>%
          filter(!is.na(responses) & is.na(test_part)) %>%
          dplyr::select(subj_id,responses) %>%
          mutate(comments = map(responses, ~fromJSON(.) %>%as.data.frame()))%>%
          unnest(comments)%>%
          dplyr::select(subj_id,debrief)%>%
          filter(!is.na(debrief) & debrief != "")%>%
          mutate(debrief = gsub("[\r\n]", "",debrief)))

@@

E4.df <- read.csv('..\\experiments\\Hangman2\\data\\jatos_results_batch2.csv',na.strings=c(""," ","NA")) %>%
  rbind(read.csv('..\\experiments\\Hangman2\\data\\jatos_results_batch1.csv',na.strings=c(""," ","NA"))) %>%
  mutate(subj_id = paste(as.character(participant_number),substr(PROLIFIC_PID,1,5),sep='')) %>%
  mutate(subj_id = factor(subj_id)) %>%
  # the dash is breaking r
  mutate(test_part = ifelse(test_part=='non-pretend','nonpretend',test_part)) %>%
  filter(PROLIFIC_PID != '5ec4a156bc5aac3819ac52f2' & #reported some of the letters not showing up on their screen
        PROLIFIC_PID != '58211fc787f6b90001f13f9' & #reported some of the letters not showing up on their screen
        PROLIFIC_PID != '6149256f6335b06ade3723e0' &  # due a PROLFIIC bug, participated twice
        PROLIFIC_PID != '615ddab1e4f013092538b6c5')# due a PROLFIIC bug, participated twice

@@

```
Similar to Battleship, the within-subject standard deviation in number of misses in non-pretend games (`r E4.sd_num_misses$nonpretend%>%mean()%>%printnum()`) was on average significantly higher than in pretend games (mean SD = `r E4.sd_num_misses$pretend%>%mean()%>%printnum()`, `r apa_print(E4.sd_num_misses$diff%>%t.test())$statistic`). Here too, pretend games were significantly more similar to each other, in terms of number of misses, than non-pretend games.

@@

E4.debrief_labeled <- read.csv('..\\experiments\\Hangman2\\data\\debrief.csv',na.strings=c(""," ","NA"))

@@

>>>>

classoption       : "man"
output            : word_document

bibliography: references.bib
---

@@

groundhog.library(c(
  'png',
  'grid',
  'ggplot2',
  'svglite',
  'xtable',
  'papaja',
  'tidyverse',
  'broom',
  'cowplot',
  'MESS', # for AUCs
  'lsr', # for effect sizes
  'pwr', # for power calculations
  'brms', # for mixed effects modeling
  'BayesFactor', # for Bayesian t test
  'jsonlite', # parsing data from sort_trial
  'caret', #for cross validation
  'ggrepel' #for word scatterplots
), '2025-06-01')

@@

p<- E2.click_df %>%
  mutate(test_part=factor(test_part,labels=c('non-pretend','pretend')))%>%
  filter(!(subj_id %in% E2.lucky_pretenders))%>%
  group_by(subj_id,test_part, genuine_first) %>%
  summarise(sd_num_clicks=sd(num_clicks)) %>%
  group_by(test_part)%>%
  summarise(
    mean=mean(sd_num_clicks),
    se=se(sd_num_clicks)
    ) %>%
  ggplot(aes(x=test_part,
             color=test_part,
             y=mean))+
  geom_point(size=6)+
  geom_errorbar(aes(ymin=mean-se,ymax=mean+se),color='black', width=0.3)+
  scale_color_manual(values=c("#69b3a2", "#404080")) +
  labs(x="",y="")+
  theme(legend.pos='na')+
  theme_classic()+
  scale_y_continuous(limits=c(0,2.8))+
  theme(legend.pos='na')

ggsave('figures/Battleship2/num_guess_sd.png',p,width=2.5,height=1.5,dpi=300)

@@

E2.debrief_labeled <- read.csv('../experiments/Battleships2/data/debrief.csv',na.strings=c(""," ","NA")) %>%
  dplyr::select(-subj_id) %>%
  merge(E2.debrief <- E2.df %>%
          filter(!is.na(responses) & is.na(test_part)) %>%
          dplyr::select(subj_id,responses) %>%
          mutate(comments = map(responses, ~fromJSON(.) %>%as.data.frame()))%>%
          unnest(comments)%>%
          dplyr::select(subj_id,debrief)%>%
          filter(!is.na(debrief) & debrief != "")%>%
          mutate(debrief = gsub("[\r\n]", "",debrief)))

@@

E4.df <- read.csv('../experiments/Hangman2/data/jatos_results_batch2.csv',na.strings=c(""," ","NA")) %>%
  rbind(read.csv('../experiments/Hangman2/data/jatos_results_batch1.csv',na.strings=c(""," ","NA"))) %>%
  mutate(subj_id = paste(as.character(participant_number),substr(PROLIFIC_PID,1,5),sep='')) %>%
  mutate(subj_id = factor(subj_id)) %>%
  # the dash is breaking r
  mutate(test_part = ifelse(test_part=='non-pretend','nonpretend',test_part)) %>%
  filter(PROLIFIC_PID != '5ec4a156bc5aac3819ac52f2' & #reported some of the letters not showing up on their screen
        PROLIFIC_PID != '58211fc787f6b90001f13f9' & #reported some of the letters not showing up on their screen
        PROLIFIC_PID != '6149256f6335b06ade3723e0' &  # due a PROLFIIC bug, participated twice
        PROLIFIC_PID != '615ddab1e4f013092538b6c5')# due a PROLFIIC bug, participated twice

@@

p<- E4.click_df %>%
  mutate(test_part=factor(test_part,levels=c('nonpretend','pretend'),labels=c('non-pretend','pretend')))%>%
  group_by(subj_id,test_part, genuine_first) %>%
  summarise(sd_num_clicks=sd(num_misses)) %>%
  group_by(test_part)%>%
  summarise(
    mean=mean(sd_num_clicks),
    se=se(sd_num_clicks)
    ) %>%
  ggplot(aes(x=test_part,
             color=test_part,
             y=mean))+
  geom_point(size=6)+
  geom_errorbar(aes(ymin=mean-se,ymax=mean+se),color='black', width=0.3)+
  scale_color_manual(values=c("#69b3a2", "#404080")) +
  labs(x="",y="")+
  theme(legend.pos='na')+
  theme_classic()+
  scale_y_continuous(limits=c(0,2.8))+
  theme(legend.pos='na')

ggsave('figures/Hangman2/num_guess_sd.png',p,width=2.5,height=1.5,dpi=300)

Similar to Battleship, the within-subject standard deviation in number of misses in non-pretend games (`r E4.sd_num_misses$nonpretend%>%mean()%>%printnum()`) was on average significantly higher than in pretend games (mean SD = `r E4.sd_num_misses$pretend%>%mean()%>%printnum()`, `r apa_print(E4.sd_num_misses$diff%>%t.test())$statistic`). Here too, pretend games were significantly more similar to each other, in terms of number of misses, than non-pretend games.

@@

E4.debrief_labeled <- read.csv('../experiments/Hangman2/data/debrief.csv',na.strings=c(""," ","NA"))

@@

### The letter frequency heuristic

In this analysis, we asked whether pretenders are doing more than guessing frequent letters such as vowels. We therefore compared behaviour in pretend and non-pretend games to the behaviour of an agent that selects the most frequent letter among the ones that were not selected so far. By focusing our analysis on misses only (that is, selections of letters that are not in the target word), we ensured that any observed advantage in real players' behaviour is not simply due to making more correct guesses.

```{r hangman-letter-freuqency}

E4.misses_with_simulations <- E4.click_log_with_boards %>%
  filter(!hit_bin)%>%
  rbind(E4.click_log_with_boards_random %>%
          filter(!hit_bin)) %>%
  rbind(E4.click_log_with_boards_most_frequent %>%
          filter(!hit_bin))

E4.nic <- E4.misses_with_simulations%>%
  group_by(subj_id,test_part, genuine_first) %>%
  summarise(nic=sum(p_click>0)/n()*100)

E4.nic %>%
  group_by(test_part) %>%
  mutate(test_part=factor(test_part,
                          levels=c('random','nonpretend','pretend','most_frequent'),
                          labels=c('random','non-pretend','pretend','most frequent'))) %>%
  summarise(p_irrational=mean(nic),
            sep_irrational=se(nic)) %>%
  ggplot(aes(x=test_part,y=p_irrational, fill=test_part))+
  geom_bar(stat="identity", color="black",
           position=position_dodge()) +
  geom_errorbar(aes(ymin=p_irrational-sep_irrational, ymax=p_irrational+sep_irrational), width=.2,
                position=position_dodge(.9)) +
  labs(x='',y='% errors that make sense') +
  scale_fill_manual(values=c("#FDE725", "#69b3a2", "#404080", "#E88567")) +
  theme_classic()+
  theme(legend.position='na')

ggsave('figures/forPresentation/hangman_rational.png',width=5,height=3,dpi=300)

Our dependent measure was the proportion of guesses that "make sense", that is, that are consistent with the board state and the word category. Pretend guesses made more sense according to this measure than what we would expect if people were only guessing the most frequent letters (difference in percentage of sensible errors between pretend and heuristic behaviour: `r apa_print(E4.nic %>% spread(test_part,nic) %>% mutate(diff=pretend- most_frequent)%>%pull(diff)%>%t.test())$full_result`).
